\name: A flexible and data-driven K-process model for nucleosynthesis

### 2023-07-10 Paper/main.tex

% to-do items
% -----------
% - Look at all latex warnings on the build

@@

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
%\received{March 1, 2021}
%\revised{April 1, 2021}
%\accepted{\today}

%\submitjournal{ApJ}

@@

\title{\name: A flexible and data-driven $K$-process model for nucleosynthesis}

@@

\author[0000-0003-2866-9403]{David W. Hogg}
\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726~Broadway, New~York,~NY 10003, USA}
\affiliation{Max-Planck-Institut f{\"u}r Astronomie, K{\"o}nigstuhl 17, D-69117 Heidelberg, Germany}
\affiliation{Flatiron Institute, 162 Fifth Avenue, New~York,~NY 10010, USA}

\author{Julianne Dalcanton}
\author{Sten Hasselquist}
\author{Bridget Ratcilffe}
\author{Melissa Ness}
\author{David Weinberg}

\begin{abstract}\noindent % trust
\add{The origin of the surface abundance patterns of Milky Way stellar populations can be explained by two dominant processes,} one prompt and one delayed.
\add{While two components can well describe the stellar abundance trends, elements are known to be produced by more than two sources.}
We analyze the abundances of 14 elements relative to Mg for 48,659 stars from APOGEE DR17 with a flexible, data-driven $K$-process model---dubbed \name.
In the fiducial model, where $K=2$, each elemental abundance of each star is described as the vector sum of a prompt and delayed process.
%While prior work derives the relative importance of \add{the prompt and delayed processes in the production of various elements} from median observed abundance trends, we relax these assumptions and 
%We simultaneously fit all model parameters to the data, with only minimal constraints to keep the processes interpretable.
We find that our flexible method is able to recover the observed abundance bimodality and that this bimodality is visible to higher [Mg/H] in process amplitude space than in abundance space.
We compare fit parameters and statistics from this new method to those from \citet{weinberg2022}, finding that the two methods produce similar results, but that \name{} better predicts stellar abundances, especially for C+N and Mn.
%We test the validity of the standard two-process model assumptions, such as that in the sun Fe is generated by a roughly half-half combination of prompt and delayed sources.
%We find that the assumptions of prior works are generally confirmed, but that the inferred fractional contributions from each process are strongly dependent upon Fe assumptions.
\add{We explore the impact of our assumptions about the origin of Fe, needed to break the rotational symmetry of the model, constraining the parameter space that is able to reproduce the observed abundance distribution and finding that the inferred fractional contribution from each process are dependent upon our choice.}
Finally, we add a third and fourth process, calibrated to Ce and Mn, respectively. The additional processes \add{give the model more freedom} and improve the model's ability to predict the stellar abundances, \add{as expected}, but demonstrates the model's ability to extend beyond $K=2$. 
\name{} enables future research to study the disk's abundance dimensionallity, investigate age-abundance label correlations, fit non-bimodal abundance distributions with a multi-component nucleosynthetic model, and much more. 
\ejg{Hogg: try to shorten the abstract. I made some changes but still need to cut like 25 more words.} 
\end{abstract}

\keywords{foo --- bar}

@@

At the present day, stellar surface abundances are not very well explained by purely \textsl{ab initio}, physics-driven models.
Theoretical yields vary from set to set, as they are dependent on progenitor properties and explosion assumptions \citep[e.g.,][]{rybizki2017, blancato2019, buck2021, griffith2021b}. 
The wide parameter space of progenitor and supernova models coupled with uncertainties in reaction rates and explosion physics hinder the creation of an accurate nucleosynthetic model from theory alone.
In the long run, it is incumbent upon us to understand these issues and correct the assumptions or calculations underlying our nucleosynthetic and astrophysical models.
In the short run, however, we gather data---tens of millions of abundance measurements on millions of stars in different astronomical surveys such as RAVE, SEGUE, LAMOST, Gaia-ESO, APOGEE/MWM, GALAH, and H3 \citep{steinmetz2006, yanny2009, gilmore2012, desilva2015, luo2015, majewski2017, conroy2019}.
This raises the question: \emph{Can we take a data-driven approach to nucleosynthesis?}

@@

We build upon the work of \citet[][hereafter G22]{griffith2019, griffith2022} and \citet[][hereafter W19, W22]{weinberg2019, weinberg2022}, who used the bimodality in [Mg/Fe] vs. [Fe/H]\footnote{Where $[{\rm X}/{\rm Y}] = \log({\rm X}/{\rm Y}) - \log({\rm X}/{\rm Y})_{\odot}$ and $({\rm X}/{\rm Y})_{\odot}$ is the solar abundance ratio.} \citep[e.g.,][]{fuhrmann1998, bensby2003, adibekyan2012} to separate stars into populations with high and low SNIa enrichment. Using the median [X/Mg] vs. [Mg/H] abundance trends, these works explain data from the GALAH\footnote{GALAH = GALactic Archaeology with HERMES.} and SDSS-IV APOGEE\footnote{APOGEE = Apache Point Observatory Galactic Evolution Experiment, part of the Sloan Digital Sky Survey} surveys, respectively, with a two-process model. Because the median abundance trends in [X/Mg] vs. [Mg/H] space are largely insensitive to aspects of chemical evolution, such as outflows and variations in star formation history \citep{weinberg2019}, the population abundance trends are set by the nucleosynthetic processes and can be used to empirically constrain Galactic enrichment.

These works, as well as \citet{ting2022} and \citet{ratcliffe2023}, find that the Milky Way stellar abundances are well fit by two components, \add{grounded in [Fe/H] and [Mg/Fe],} down to residuals of 0.01 to 0.02 dex. Simultaneously, \citet{frankel2018} and \citet{ness2022} have found that disk abundances are also well described by a two-component model of birth radius and age. Correlations between two-process model parameters and stellar ages and kinematics (W22) \add{as well as the success of a two-component model of [Fe/H] and age in predicting APOGEE abundances \citep{ness2019}} suggest that these two 2-dimensional models are somehow interconnected.

Beyond standard CCSN and SNIa enrichment, many elements have contributions from additional nucleosynthetic processes, such as the rapid ($r$) and slow ($s$) neutron capture processes \citep[e.g.,][]{arlandini1999, bisterzo2014} in asymptotic giant branch (AGB) stars \citep[e.g.,][]{simmerer2004, karakas2016} and merging neutron stars \citep[e.g.,][]{kilpatrick2017}, or atypical supernova explosions \citep[e.g.,][]{nomoto2013}. After predicting stellar abundances from $\feh$ and $\mgfe$, \citet{ting2022} identify correlated abundance residuals that are unexplained by observational uncertainties, indicative of additional nucleosynthetic processes that standard disk CCSN and SNIa enrichment cannot explain. Results from G22 and W22 support this conclusion, and both works attempt to add additional processes to their models to account for non-CCSN and non-SNIa enrichment, though in a very restrictive manner. Other sources of abundance scatter, such as stochastic sampling of the Initial Mass Function (IMF), IMF variations, and bursty star formation history could also cause deviations away from a two-process model \citep{belokurov2018, griffith2023}.

@@

In this work, our main innovations are to relax the assumptions made in G22 and W22, to be more agnostic about the nucleosynthetic processes, and to be more principled with the measurements or inferences from data.
In the $K$ Process Model (\name{}), we find the intersection between reliable facts about nucleosynthesis and good abundance measurements to build an edifice of Galactic enrichment.
The model is hierarchical, in that it learns some parameters (process vectors) that are shared across all stars, but different for each element, and some parameters (process amplitudes) that are shared across all elements, but different for each star.
The parameters output by our model can thus be used as de-noised abundance labels; these will sharpen relationships between abundances and stellar parameters (including birth location and time). \add{Our main contribution is to construct a data-driven model for nucleosynthesis that has good statistical properties, only enforcing constraints to break rotational degeneracies. All other \name{} parameters are set by the data with no fixed normalization.}
% Had said some of this in the sentences before...
%The process parameters are derived Uses a hierarchical statistical model so it combines abundances into K process amplitudes for each star and denoised predicted abundances of each star, and metallicity-dependent effective yields for the processes.

@@

As in W22 and G22, we propose that all stellar abundances can be generated by a combination of $K$ nucleosynthetic processes.
In this picture, each element has $K$ metallicity-dependent process vector components that are shared across the full stellar sample, while each star individually has $K$ process amplitudes, \add{which apply across all elements, such that the expected logarithmic abundance of element $j$ in star $i$ ($m_{ij}$) is defined as:}
\begin{equation}\label{eq:mij_k}
    m_{ij} = \log_{10} \, \sum^K_{k=1} A_i^k \, q_{k,j}^Z ~.
\end{equation}
Each star $i$ has $K$ process amplitudes ($A^k_i$)
and each element $j$ has $K$ \add{metallicity dependent} process vector components ($q_{k,j}^{Z}$). 
\add{The $Z$ superscript denotes the dependence of the process vectors on metallicity, $Z$, taken to be [Mg/H].}
The $m_{ij}$ value is the model expected logarithmic abundance of element $j$ \add{relative to Hydrogen in star $i$, such that the observed abundance can be expressed}
\begin{equation}\label{eq:xh}
    [\text{X}_j/\text{H}]_i = m_{ij} + \text{noise},
\end{equation}
\add{where ``noise'' represents observational noise and/or other sources of intrinsic abundance scatter that are not included in this model.}

\add{For example, in the $K=2$ case for a star ($i=40$) with a O abundance of $[\text{O}/\text{H}] =0.155$ and $\mgh = 0.122$, \name{} fits $q_{\text{CC,O}}^{\mgh=0.12} = 0.91$ and $q_{\text{Ia,O}}^{\mgh=0.12} = 0.13$. This indicates that at $\mgh = 0.12$, stellar O is dominantly produced by the prompt (CCSN) process. For this star, \name{} also derives $A_{40}^{\rm CC} = 1.3$ and $A_{40}^{\rm Ia} = 1.2$ from the abundances of all 15 elements. Summing the products of the corresponding process vector component and process amplitude, we find $m_{40,\text{O}} = \log_{10} ( 1.36 ) = 0.135$. \name{} predicts the star's O abundance to 0.02 dex, the typical scale of observational noise on [O/H] in APOGEE. For more detailed examples from a similar model, see Section 2 of W22.}
\ejg{Hogg let me know what you think. We don't introduce the CC and Ia notation until later in this section so I think if it stays it needs more clarification. I lean towards just sending people to Section 2 and Figures 2 and 3 of W22. }

@@

\paragraph{1. $K$ processes}
All elements on the periodic table are produced by a combination of nucleosynthetic processes such as CCSN, SNIa, AGB stars, and merging neutron stars \citep{johnsonja2020}. The majority of $\alpha$, light odd-Z, and Fe-peak elements (the elements observed by APOGEE) are dominantly produced by $K=2$ sources, with one being a prompt process or mix of prompt processes, and one being a delayed process or a mix of delayed processes. This is substantiated by theoretical yields \citep[e.g.,][]{anderson2019, rybizki2017} and past successful data-driven models (e.g., \citealp{ness2019}, G22, W22, \citealp{ting2022, ratcliffe2023}). In this paper we therefore assume that $K \geq 2$, though \name{} could in principle be implemented with $K=1$.

\paragraph{2. Linearity}
At every metallicity, the (linear) $(X/H)$ abundances of a star can be expressed as a linear combination of $K$ processes.
These processes themselves will depend on metallicity, but a linear sum is sufficient to explain all element abundances at any overall metallicity.
Because yields depend on stellar masses and progenitor structure, and because different stars can get to their metallicities by different histories, this assumption must be at least slightly wrong in detail.

@@

In \name{}, this assumption is enforced by fixing the Mg process vector components such that
\begin{equation}\label{eq:qcc_mg_solar}
    q_{{\rm CC, Mg}}^{\,Z} = 1, \quad q_{k>1,{\rm Mg}}^{\,Z} = 0
\end{equation}
at all metallicities.

@@

\begin{equation}\label{eq:qcc_fe_solar}
    q_{{\rm CC, Fe}}^{\,Z} = 0.4, \quad q_{{\rm Ia, Fe}}^{\,Z} = 1 - q_{{\rm CC, Fe}}^{\,Z} \quad q_{k>2,{\rm Fe}}^{\,Z} = 0
\end{equation}
at all metallicities. 
This assumption places a star with purely prompt enrichment on the low-metallicity $\mgfe$ plateau near 0.4 dex, in agreement with APOGEE observations but in contention with recent results from \citet{conroy2022} which place the $\mgfe$ plateau near 0.6 dex.
We explore the impact of different $\qccFe$ assumptions in Section~\ref{subsec:qccFe}.

@@

\paragraph{7. APOGEE abundances and uncertainties}
We assume that the APOGEE abundances and uncertainties can be used for this project.
This is not the same as assuming that they are correct, but rather that it is possible and useful to build an interpretable model to explain them. We describe the potential data systematics in Section~\ref{sec:data}.
\add{For our purposes, we care merely about the statistical observational errors rather than systematics that arise from imperfect modeling of the spectra such as NLTE effects, though differential systematics across the sample can artificially add abundance scatter. The actual derived values of $q_{k,j}^{Z}$ will be affected by systematic offsets in the abundances. We add a softening parameter $Q$ (Equation~\ref{eq:inflate_ivar}) to allow for the possibility that APOGEE observational errors are underestimated or that there is intrinsic scatter around the \name{} predictions.}
% The APOGEE errors may be slightly underestimated for some element measurements, so we add a softening parameter $Q$ to inflate the uncertainties on abundance-outlier stars (see Equations~\ref{eq:chi2} and ~\ref{eq:inflate_ivar}).

\paragraph{8. Robust likelihood function}
The observed value of [X/H] can be described as the $K$ process expected value plus observational noise and/or other sources of intrinsic abundance scatter, \add{as described by Equation~\ref{eq:xh}.}
The expression in this equation can be thought of as the key assumption underlying our likelihood function.
In detail the (negative two times the) log likelihood function is given by a chi-squared ($\chi^2$) objective
\begin{equation}\label{eq:chi2}
    \chi^2 = \sum_{ij}\frac{1}{\sigma_{ij}^2} \, (\xh_{ij} - m_{ij})^2 ~,
\end{equation}
where $1/\sigma_{ij}^2$ is the (robust; see below) inverse variance on measurement $ij$.

Because we don't want to be too drawn or influenced by outlier points, we don't use the observed errors $\sigma_{{\rm obs},ij}$ in the likelihood, but instead we soften them in the spirit of iteratively reweighted least squares (\citealt{irls}):
\begin{equation}\label{eq:inflate_ivar}
    \frac{1}{\sigma_{ij}^2} = \frac{Q^2/\sigma_{{\rm obs},ij}^2}{Q^2 + (\xh_{ij} - m_{ij})^2 /\sigma_{{\rm obs},ij}^2 },    
\end{equation}
where $Q$ is a softening parameter. \add{Our results are largely insensitive to the choice of $Q$ (e.g., $1 < Q < 10$), so we choose to set $Q=5$. Very small values (e.g., $Q=.1$) will erase some of the abundance structure and produce poorer fits.}

\paragraph{9. Implementation and optimization} 
With the above assumptions in place, the likelihood function can be optimized to a set of stellar abundances.
The model is initialized at the Mg and Fe process vector components from Equations~\ref{eq:qcc_mg_solar} --~\ref{eq:qcc_fe_z}. It subsequently optimizes the process amplitudes (dubbed the $A$-step) using only Mg and Fe at fixed process vector components, and then optimizes the process vector components (dubbed the $q$-step) for all elements at fixed process amplitudes.
The $A$-step and $q$-step are alternated, repeating 48 rounds of optimization, \add{in the $K=2$ case,} and updating the best-fit parameters when the objective function improves. We find few differences in the best-fit parameters when we decrease the number of iterations to 32, indicating that the model quickly finds a good solution.  
In detail the optimizations are performed with a nonlinear $\chi^2$ minimization algorithm (Gauss--Newton nonlinear least-squares) from \texttt{jaxopt}\footnote{\url{https://jaxopt.github.io/}}.

@@

APOGEE DR17 reports stellar parameters, including $\teff$ and $\logg$, as well as 20 elemental abundances: C, \ion{C}{1}, N, O, Na, Mg, Al, Si, S, K, Ca, \ion{Ti}{1}, \ion{Ti}{2}, V, Cr, Mn, Fe, Co, Ni, and Ce for 657,135 stars. Among these elements and ions, some are measured more precisely than others. We exclude Ti from our analysis as there are large differences between the abundances derived from the \ion{Ti}{1} and \ion{Ti}{2} lines \citep{jonsson2020}. We also exclude P and V, as the P abundances are measured from a few very weak spectra features and V abundances are one of the least precise and least accurate labels \citep{jonsson2020}. Both P and V display strong abundance artifacts and large scatter. Among the remaining elements, we note the following concerns: weak Na spectral features, large abundance scatter in S, significant systematic artifacts in Cr abundances at super-solar metallicities, potentially strong unaccounted-for NLTE effects on Mn abundances \citep{bergemann2019}, and large abundance scatter in Co and Ce---both derived from one line. Further, we note that DR17 is the first time Ce abundances have been included in APOGEE data products. \ejg{Comments that this is not correct and more Ce lines are used, but the DR17 paper isn't out yet. Sten to help.} For a more detailed discussion of abundance systematics and their effects on population trends, see \citet{jonsson2020} and \citet{griffith2021a}.

@@

These cuts result in a sample of 48,659 stars \add{that span the Galactic disk. We plot their $Z$ vs. $R$ distribution, as well as the distributions of distances and eccentricities in Figure~\ref{fig:star_dist}, taking distances and kinematics from \citep{queiroz2023}. While our stellar sample extends from Galactic center, to the outer disk, to the halo, the majority of our stars ($75\%$) are within 3.5 kpc of the sun. Further, $94\%$ of our stellar sample has an eccentricity less than 0.4, indicative of \textit{in situ} origin. In this paper, we assume that the KPM fits will be consistent across the Galactic disk, as the median high-Ia and low-Ia [X/Mg] vs. [Mg/H] abundance tends are insensitive to Galactic location (W19, \citealp{griffith2021a}.}   
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/star_sample.pdf}
    \caption{\add{Left: distribution of our stellar sample in $Z$ (kpc) vs. $R$ (kpc). Center: distribution of distances (kpc). Right: distribution of eccentricity. Our stellar sample spans the Galactic disk, but the majority of our stars are within 3.5 kpc of the sun.}}
    \label{fig:star_dist}
\end{figure*}

>>>>

% to-do items
% -----------

@@

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
% \received{July 10, 2023}
% \revised{}
% \accepted{}

\submitjournal{ApJ}

@@

\title{\name: A flexible and data-driven K-process model for nucleosynthesis}

@@

\author[0000-0003-2866-9403]{David W. Hogg}
\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726~Broadway, New~York,~NY 10003, USA}
\affiliation{Max-Planck-Institut f{\"u}r Astronomie, K{\"o}nigstuhl 17, D-69117 Heidelberg, Germany}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Avenue, New York, NY 10010, USA}

\author{Julianne J. Dalcanton}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Avenue, New York, NY 10010, USA}
\affiliation{Department of Astronomy, Box 351580, University of Washington, Seattle, WA 98195}

\author[0000-0001-5388-0994]{Sten Hasselquist}
\affiliation{Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD 21218}

\author[0000-0003-1124-7378]{Bridget Ratcilffe}
\affiliation{Leibniz-Institut f\"{u}r Astrophysik Potsdam (AIP), An der Sternwarte 16, 
14482 Potsdam, Germany}

\author[0000-0001-5082-6693]{Melissa Ness}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Avenue, New York, NY 10010, USA}
\affiliation{Department of Astronomy, Columbia University, Pupin Physics Laboratories, New York, NY 10027, USA}

\author[0000-0001-7775-7261]{David H. Weinberg}
\affiliation{The Department of Astronomy and Center of Cosmology and AstroParticle Physics, The Ohio State University, Columbus, OH 43210, USA}

\begin{abstract}\noindent
The element abundance pattern found by APOGEE and GALAH in Milky Way disk stars is close to two-dimensional, dominated by production from one prompt process and one delayed process.
This simplicity is remarkable, since the elements are produced by a multitude of nucleosynthesis mechanisms operating in stars with a wide range of progenitor masses.
We fit the abundances of 14 elements for 48,659 red-giant stars from APOGEE DR17 using a flexible, data-driven $K$-process model---dubbed \name.
In our fiducial model, with $K=2$, each abundance in each star is described as the sum of a prompt and a delayed process contribution.
We find that \name{} with $K=2$ is able to explain the abundances well, recover the observed abundance bimodality, and detect the bimodality over a greater range in metallicity than previously has been possible.
We compare to prior work by \citet{weinberg2022}, finding that \name{} produces similar results, but that \name{} better predicts stellar abundances, especially for elements C+N and Mn and for stars at super-solar metallicities.
The model makes assumptions, including especially that it fixes some parameters to break degeneracies and improve interpretability; we find that some of the nucleosynthetic implications are dependent upon these detailed parameter choices.
We add a third and fourth process (to make $K=4$), finding that the additional processes give the model more freedom and improve the model's ability to predict the stellar abundances, as expected, but they don't qualitatively change the story. 
The results of \name{} have implications for the formation of the Galaxy disk, the relationship between abundances and ages, and the physics of nucleosynthesis.
\end{abstract}

\keywords{Nucleosynthesis (1131), Stellar abundances (1577), Galaxy chemical evolution (580), Core-collapse supernovae (304), Type Ia supernovae (1728), }

@@

At the present day, stellar surface abundances are not very well explained by purely \textsl{ab initio}, physics-driven models.
Theoretical yields vary from data set to data set, as they are dependent on progenitor properties and explosion assumptions \citep[e.g.,][]{rybizki2017, blancato2019, buck2021, griffith2021b}. 
The wide parameter space of progenitor and supernova models coupled with uncertainties in reaction rates and explosion physics hinder the creation of an accurate nucleosynthetic model from theory alone.
In the long run, it is incumbent upon us to understand these issues and correct the assumptions or calculations underlying our nucleosynthetic and astrophysical models.
In the short run, however, we gather data---tens of millions of abundance measurements on millions of stars in different astronomical surveys such as RAVE, SEGUE, LAMOST, Gaia-ESO, APOGEE/MWM, GALAH, and H3 \citep{steinmetz2006, yanny2009, gilmore2012, desilva2015, luo2015, majewski2017, conroy2019}.
This raises the question: \emph{Can we take a data-driven approach to nucleosynthesis?}

@@

We build upon the work of \citet[][hereafter G22]{griffith2019, griffith2022} and \citet[][hereafter W19, W22]{weinberg2019, weinberg2022}, who used the bimodality in [Mg/Fe] vs. [Fe/H]\footnote{Where $[{\rm X}/{\rm Y}] = \log({\rm X}/{\rm Y}) - \log({\rm X}/{\rm Y})_{\odot}$ and $({\rm X}/{\rm Y})_{\odot}$ is the solar abundance ratio.} \citep[e.g.,][]{fuhrmann1998, bensby2003, adibekyan2012} to separate stars into populations with high and low SNIa enrichment. Using the median [X/Mg] vs. [Mg/H] abundance trends, these works explain data from the GALAH\footnote{GALAH = GALactic Archaeology with HERMES.} and SDSS-IV APOGEE\footnote{APOGEE = Apache Point Observatory Galactic Evolution Experiment, part of the Sloan Digital Sky Survey} surveys, respectively, with a two-process model. Because the median abundance trends in [X/Mg] vs. [Mg/H] space are largely insensitive to aspects of chemical evolution, such as outflows and variations in star formation history (W19), the population abundance trends are set by the nucleosynthetic processes and can be used to empirically constrain Galactic enrichment.

These works, as well as \citet{ting2022} and \citet{ratcliffe2023}, find that the Milky Way stellar abundances are well fit by two components, grounded in [Fe/H] and [Mg/Fe], down to residuals of 0.01 to 0.03 dex for the most precisely measured elements and 0.05 to 0.1 dex for elements (such as Na, C, and Ce) with large measurement errors. Simultaneously, \citet{frankel2018} and \citet{ness2022} have found that disk abundances are also well described by a two-component model of birth radius and age. Correlations between two-process model parameters and stellar ages and kinematics (W22) as well as the success of a two-component model of [Fe/H] and age in predicting APOGEE abundances \citep{ness2019} suggest that these two 2-dimensional models are somehow interconnected.

Beyond standard CCSN and SNIa enrichment, many elements have contributions from additional nucleosynthetic processes, such as the rapid ($r$) and slow ($s$) neutron capture processes \citep[e.g.,][]{arlandini1999, bisterzo2014} in asymptotic giant branch (AGB) stars \citep[e.g.,][]{simmerer2004, karakas2016}, merging neutron stars \citep[e.g.,][]{kilpatrick2017}, or atypical supernova explosions \citep[e.g.,][]{nomoto2013}. After predicting stellar abundances from $\feh$ and $\mgfe$, \citet{ting2022} identify correlated abundance residuals that are unexplained by observational uncertainties, indicative of additional nucleosynthetic processes that standard disk CCSN and SNIa enrichment cannot explain. Results from G22 and W22 support this conclusion, and both works attempt to add additional processes to their models to account for non-CCSN and non-SNIa enrichment, though in a restrictive manner. Other sources of abundance scatter, such as stochastic sampling of the Initial Mass Function (IMF), IMF variations, and bursty star formation history could also cause deviations away from a two-process model \citep{belokurov2018, griffith2023}.

@@

In this work, our main innovations are to relax the assumptions made in G22 and W22, to be more agnostic about the nucleosynthetic processes, and to be more principled with the measurements or inferences from data.
In the $K$ Process Model (\name{}), we find the intersection between reliable facts about nucleosynthesis and good abundance measurements to build an edifice of Galactic enrichment.
The model is hierarchical, in that it learns some parameters (process vectors) that are shared across all stars, but different for each element, and some parameters (process amplitudes) that are shared across all elements, but different for each star.
The parameters output by our model can thus be used as de-noised abundance labels; these will sharpen relationships between abundances and stellar parameters (including birth location and time). Our main contribution is to construct a data-driven model for nucleosynthesis that has good statistical properties, only enforcing constraints to break rotational degeneracies. All other \name{} parameters are set by the data with no fixed normalization.

@@

As in W22 and G22, we propose that all stellar abundances can be generated by a combination of $K$ nucleosynthetic processes.
In this picture, each element has $K$ metallicity-dependent process vector components that are shared across the full stellar sample, while each star individually has $K$ process amplitudes, which apply across all elements, such that the expected logarithmic abundance of element $j$ relative to H in star $i$ ($m_{ij}$) is defined as:
\begin{equation}\label{eq:mij_k}
    m_{ij} = \log_{10} \, \sum^K_{k=1} A_i^k \, q_{k,j}^Z ~.
\end{equation}
Each star $i$ has $K$ process amplitudes ($A^k_i$)
and each element $j$ has $K$ metallicity dependent process vector components ($q_{k,j}^{Z}$). 
The $Z$ superscript denotes the dependence of the process vectors on metallicity, $Z$, taken to be [Mg/H].
The observed abundance can be expressed
\begin{equation}\label{eq:xh}
    [\text{X}_j/\text{H}]_i = m_{ij} + \text{noise},
\end{equation}
where ``noise'' represents observational noise and/or other sources of intrinsic abundance scatter that are not included in this model. For detailed examples of a similar model with $K=2$, see Section 2 and Figures 2 and 3 of W22.

@@

\paragraph{1. $K$ processes}
All elements on the periodic table are produced by a combination of nucleosynthetic processes such as CCSN, SNIa, AGB stars, and merging neutron stars \citep{johnsonja2020}. The majority of $\alpha$, light odd-$Z$, and Fe-peak elements (the elements observed by APOGEE) are dominantly produced by $K=2$ sources, with one being a prompt process or mix of prompt processes, and one being a delayed process or a mix of delayed processes. This is substantiated by theoretical yields \citep[e.g.,][]{anderson2019, rybizki2017} and past successful data-driven models (e.g., \citealp{ness2019}, G22, W22, \citealp{ting2022, ratcliffe2023}). In this paper we therefore assume that $K \geq 2$, though \name{} could in principle be implemented with $K=1$.

\paragraph{2. Linearity}
At every metallicity, the (linear) (X/H) abundances of a star can be expressed as a linear combination of $K$ processes.
These processes themselves will depend on metallicity, but a linear sum is sufficient to explain all element abundances at any overall metallicity.
Because yields depend on stellar masses and progenitor structure, and because different stars can get to their metallicities by different histories, this assumption must be at least slightly wrong in detail.

@@

In \name{}, this assumption is enforced by fixing the Mg process vector components such that
\begin{equation}\label{eq:qcc_mg_solar}
    q_{{\rm CC, Mg}}^{\,Z} = 1, \quad q_{k>1,{\rm Mg}}^{\,Z} = 0
\end{equation}
at all metallicities. Equation~\ref{eq:qcc_mg_solar} also imposes that the Mg process is metallicity independent.

@@

\begin{equation}\label{eq:qcc_fe_solar}
    q_{{\rm CC, Fe}}^{\,Z} = 0.4, \quad q_{{\rm Ia, Fe}}^{\,Z} = 1 - q_{{\rm CC, Fe}}^{\,Z} \quad q_{k>2,{\rm Fe}}^{\,Z} = 0
\end{equation}
at all metallicities. 
This assumption places a star with purely prompt enrichment on the low metallicity $\mgfe$ plateau near 0.4 dex, in agreement with APOGEE observations but in contention with recent results from \citet{conroy2022} which place the $\mgfe$ plateau near 0.6 dex.
We explore the impact of different $\qccFe$ assumptions in Section~\ref{subsec:qccFe}.

@@

\paragraph{7. APOGEE abundances and uncertainties}
We assume that the APOGEE abundances and uncertainties can be used for this project.
This is not the same as assuming that they are correct, but rather that it is possible and useful to build an interpretable model to explain them. We describe the potential data systematics in Section~\ref{sec:data}.
For our purposes, we care mainly about the statistical observational errors rather than systematics that arise from imperfect modeling of the spectra such as NLTE effects, though differential systematics across the sample can artificially add abundance scatter. The actual derived values of $q_{k,j}^{Z}$ will be affected by systematic offsets in the abundances. We add a softening parameter $Q$ (Equation~\ref{eq:inflate_ivar}) to allow for the possibility that APOGEE observational errors are underestimated or that there is intrinsic scatter around the \name{} predictions.

\paragraph{8. Robust likelihood function}
The observed value of [X/H] can be described as the $K$ process expected value plus observational noise and/or other sources of intrinsic abundance scatter, as described by Equation~\ref{eq:xh}.
The expression in this equation can be thought of as the key assumption underlying our likelihood function.
In detail the (negative two times the) log likelihood function is given by a chi-squared ($\chi^2$) objective
\begin{equation}\label{eq:chi2}
    \chi^2 = \sum_{ij}\frac{1}{\sigma_{ij}^2} \, (\xh_{ij} - m_{ij})^2 ~,
\end{equation}
where $1/\sigma_{ij}^2$ is the (robust; see below) inverse variance on measurement $ij$.

Because we don't want to be too drawn or influenced by outlier points, we don't use the observed errors $\sigma_{{\rm obs},ij}$ in the likelihood, but instead we soften them in the spirit of iteratively reweighted least squares (\citealt{irls}):
\begin{equation}\label{eq:inflate_ivar}
    \frac{1}{\sigma_{ij}^2} = \frac{Q^2/\sigma_{{\rm obs},ij}^2}{Q^2 + (\xh_{ij} - m_{ij})^2 /\sigma_{{\rm obs},ij}^2 },    
\end{equation}
where $Q$ is a softening parameter. Our results are largely insensitive to the choice of $Q$ (e.g., $1 < Q < 10$), so we choose to set $Q=5$. Very small values (e.g., $Q=.1$) will erase some of the abundance structure and produce poorer fits.

\paragraph{9. Implementation and optimization} 
With the above assumptions in place, the likelihood function can be optimized to a set of stellar abundances.
The model is initialized at the Mg and Fe process vector components from Equations~\ref{eq:qcc_mg_solar} and~\ref{eq:qcc_fe_solar}. It subsequently optimizes the process amplitudes (dubbed the $A$-step) using only Mg and Fe at fixed process vector components, and then optimizes the process vector components (dubbed the $q$-step) for all elements at fixed process amplitudes.
The $A$-step and $q$-step are alternated, repeating 48 rounds of optimization, in the $K=2$ case, and updating the best-fit parameters when the objective function improves. We find few differences in the best-fit parameters when we decrease the number of iterations to 32, indicating that the model quickly finds a good solution.  
In detail the optimizations are performed with a nonlinear $\chi^2$ minimization algorithm (Gauss--Newton nonlinear least-squares) from \texttt{jaxopt}\footnote{\url{https://jaxopt.github.io/}}.

@@

APOGEE DR17 reports stellar parameters, including $\teff$ and $\logg$, as well as 20 elemental abundances: C, \ion{C}{1}, N, O, Na, Mg, Al, Si, S, K, Ca, \ion{Ti}{1}, \ion{Ti}{2}, V, Cr, Mn, Fe, Co, Ni, and Ce for 657,135 stars. In DR17, new spectral libraries \citep{hubeny2021} are generated using the Synspec code and incorporate NLTE corrections for Na, Mg, K, and Ca \citep{osorio2020}. Among the reported elements and ions, some are measured more precisely than others. We exclude Ti from our analysis as there are large differences between the abundances derived from the \ion{Ti}{1} and \ion{Ti}{2} lines \citep{jonsson2020}. We also exclude P and V, as the P abundances are measured from a few very weak spectra features and V abundances are one of the least precise and least accurate labels \citep{jonsson2020}. Both P and V display strong abundance artifacts and large scatter. Among the remaining elements, we note the following concerns: weak Na spectral features, large abundance scatter in S, significant systematic artifacts in Cr abundances at super-solar metallicities, potentially strong unaccounted-for NLTE effects on Mn abundances \citep{bergemann2019}, and large abundance scatter in Co and Ce. For a more detailed discussion of abundance systematics and their effects on population trends, see \citet{jonsson2020} and \citet{griffith2021a}.

@@

These cuts result in a sample of 48,659 stars that span the Galactic disk. We plot their $Z$ vs. $R$ location, as well as the distributions of distances and eccentricities in Figure~\ref{fig:star_dist}, taking distances and kinematics from \citep{queiroz2023}. While our stellar sample extends from Galactic center, to the outer disk, to the halo, the majority of our stars ($75\%$) are within 3.5 kpc of the sun. Further, $94\%$ of our stellar sample has an eccentricity less than 0.4, indicative of \textit{in situ} origin. In this paper, we assume that the KPM fits will be consistent across the Galactic disk, as the median high-Ia and low-Ia [X/Mg] vs. [Mg/H] abundance tends are insensitive to Galactic location (W19, \citealp{griffith2021a}).  
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/star_sample.pdf}
    \caption{Left: distribution of our stellar sample in $Z$ (kpc) vs. $R$ (kpc) where (0,0) is the Galactic center. Center: distribution of stellar distances (kpc). Right: distribution of stellar eccentricity. Our stellar sample spans the Galactic disk, but the majority of our stars are within 3.5 kpc of the sun and have kinematics consistent with \textit{in situ} origin.}
    \label{fig:star_dist}
\end{figure*}
